Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.

### 2024-01-06 apples_and_oranges.qmd
results have more logical consistency (I hope) have added html comments for Charlotte.

---
title: Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.
subtitle: Re-visiting the relative efficacy Control Conditions..
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: charlotte.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Dayna Bhudia
    email: dayna.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Lucy Foulkes
    email: lucy.foulkes@psych.ox.ac.uk
    affiliations:
        - id: OU
          name: University of Oxford
          department: Experimental Psychology
          address: Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Road
          city: Oxford
          postal-code: OX2 6GG
          country: United Kingdom
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note: 
abstract: |
  To follow
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
bibliography: emotion_concepts_paper.bib
format: 
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super

---

@@

# Method
## Studies included

NOTE: **Charlotte** to please check code lines 986 to 1121. **CB**: Done.
NOTE: **Charlotte** please check that the studies I have in the *df_long_for_metan* dataframe makes sense and correspond to what you would expect. In the beginning of the results below, I explain in detail what is missing where. **CB**: I've checked and there are fewer psy active conditions that I expected - it seems some of the trials with multiple active arms have been filtered out. I have now fixed this and the numbers look correct. The numbers look right for med studies. One thing - for psy the no of studies with missing Cohens d for controls should probably be considered 6 (not 7) as it is absent for March/TADS because we removed it (didn't want placebo to be considered as a psy control) rather than there being missing data.

## Statistical Analysis
### Measures of Effect
As the measure of effect of each individual study, we used the Standardised Mean Difference (SMD), which we defined following NOTE: **Charlotte** cite Lakens 2013 and in there Cummings 2012, as:

@@

In particular, we simulated one thousand truncated distribution of standard errors with the following general characteristics:

@@

@@

### Random Effects Metaregression
We estimated the pooled standardised mean difference for each arm by using a random effects metanalysis implemented in R's metafor package. The main underlying assumption of random effects metanalysis is that each study's true effect size $\theta_{k}$ is affected not only by sampling error $\epsilon_{k}$, but also by $\zeta$ which represents heterogeneity between studies, allowing each study's estimate to vary along a distribution of effects, and the distribution of true effect sizes termed $\tau^2$.
Therefore, we can estimate a two stage model with:

$$Y_i \sim \mathcal{N}(\theta_i, \sigma_i^2)
$$ {#eq-5}

@@

$$u_i \sim N(0, \tau^2) $${#eq-8}

@@

We used maximum likelihood (REML) to estimate model and applied  Hartung-Knapp adjustment to reduce the chance of false positives (NOTE: **Charlotte**cite Ioannides on this).

@@

The data for the studies included in this metanalysis are summarised in Supplementary Table 1 and are also available as a csv dataframe on \[<https://github.com/transatlantic-comppsych/apples_oranges>\].

@@

# add a new multilevel variable for the regression

@@

# extract coefficients and model characteristics.
condition <- levels(list_df_simulated[[1]]$four_level_var)
aggregate_results_overall <-  aggregate_model_results(list_model_1_meta_reg, condition)

@@

NOTE: CHARLOTTE Please create a table here with the results from the count studies or study count variables that you checked in the code chunk above for each A,B and C. 
The table for A should be added here and accompany the description below. 
THe other two should go below as supplementary.
As indicated in the methods we should have one overall table with all studies and all their characteristics ala Cipriani.

@@

We then applied metaregression to obtain the means and confidence intervals of each of the four arms.

We examined the variability that the simulations of the correlations of the pre-post outcomes produced. As shown in Supplemenatry Table 2, the standard deviation of the simulated mean varied in the third decimal point, indicating minimal influence of the correlations on the outcomes.

Figure 1 graphs the results with the means of each level alongside their confidence intervals. Table 1 shows the results containing the means and standard errors for each level, derived as the average of 1000 metaregressions. 
As can be intuited from the confidence intervals displayed, there are statistically significant differences at p<0.05 between the two control arms. Table 2 illustrates this with the output of one of the simulated models.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# plot results from simulated means ---------------------------------------
# 
subtitle_text_all <- "Metanalytic estimates of within-group changes: all studies"
plot_means_all <- plot_means_function(df_mean_coefs_from_sim, subtitle_text_all)
print(plot_means_all)

subtitle_text_cbt <- "Metanalytic estimates of within-group changes: cbt, fluox, escit"
plot_means_cbt <- plot_means_function(coef_and_se_means_cbt_fluox_esc_study , subtitle_text_cbt)
print(plot_means_cbt)

subtitle_text_no_wl <- "Metanalytic estimates of within-group changes: wl excluded"
plot_means_wl <- plot_means_function(coef_and_se_means_no_wl, subtitle_text_no_wl)
print(plot_means_wl)

```{r, warning=FALSE, message = FALSE, echo=FALSE}
#| label: Table 1
#| tbl-cap: Means for the four Trial Arms alongside 95% Confidence Intervals
tab_1_df <- df_mean_coefs_from_sim[, 1:length(df_mean_coefs_from_sim)-1]

knitr:: kable(tab_1_df, digits = 2)
```

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# #| label: Table 2
# #| tbl-cap: Statistics of the Difference Between Medication and Psychotherapy Control Conditions
# knitr:: kable(coefficients_from_contrasts_medcontrol_psy_control,col.names = c("Beta", "SE", "Lower CI", "Upper CI", "z-value", "p-value"), digits = 2)

We then examined whether restricting studies to the trials (US-) licensed medication fluoxetine and escitalopram and the most widely used and evidence-based psychological treatment, namely CBT would change results. As can be seen, the coefficient of the difference between placebo and psychotherapy control condition is very strong as judged by the z-value of the difference (8.8), and comaprable to the overall results.

@@

stab_sims <- stability_sims %>%
  ggplot(aes(x = n_sims, y = z_value)) +
  geom_point() +
  geom_hline(yintercept = 1.65, linetype = "dotted", color = "red") +
  geom_segment(aes(x = 300, xend = 300, y = 2, yend = 1.65),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  annotate("text", x = 300, y = 2.0, label = "p < 0.05 threshold (values above line significant)", hjust = -0.1, vjust = 0) +
  geom_segment(aes(x = 300, xend = 300, y = 7.5, yend = min(stability_sims$z_value-0.3)),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  annotate("text", x = 300, y = 7.5, label = "simulated z-values",
           hjust = -0.1, vjust = 0) +
  theme_minimal() +
  labs(
    x = "Number of Simulations",
    y = "z-value",
    title = "Stability of the Statistic of the Difference between \nMedication and Psychotherapy Control Arms",
    subtitle = "Results from 1000 simulations with within-group standard errors"
  )
stab_sims

>>>>

---
title: Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.
subtitle: Re-visiting the relative efficacy Control Conditions..
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: charlotte.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Dayna Bhudia
    email: dayna.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Lucy Foulkes
    email: lucy.foulkes@psych.ox.ac.uk
    affiliations:
        - id: OU
          name: University of Oxford
          department: Experimental Psychology
          address: Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Road
          city: Oxford
          postal-code: OX2 6GG
          country: United Kingdom
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note: 
abstract: |
  To follow
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
bibliography: emotion_concepts_paper.bib
format: 
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

# Method

## Studies included

NOTE: **Charlotte** to please check code lines 986 to 1121. **CB**: Done. NOTE: **Charlotte** please check that the studies I have in the *df_long_for_metan* dataframe makes sense and correspond to what you would expect. In the beginning of the results below, I explain in detail what is missing where. **CB**: I've checked and there are fewer psy active conditions that I expected - it seems some of the trials with multiple active arms have been filtered out. I have now fixed this and the numbers look correct. The numbers look right for med studies. One thing - for psy the no of studies with missing Cohens d for controls should probably be considered 6 (not 7) as it is absent for March/TADS because we removed it (didn't want placebo to be considered as a psy control) rather than there being missing data. <!--# Charlotte, that's all great. What we now need here is an account of exactly how  we searched and how we used the other datasets. Follow previous examples, though in our case, it will be a lot about how we used the accounts of others. Please make sure to say where that we asked about missing information.  -->

## Statistical Analysis

### Measures of Effect

As the measure of effect of each individual study, we used the Standardised Mean Difference (SMD), which we defined following NOTE: **Charlotte** cite Lakens 2013 and in there Cummings 2012, as:

@@

In particular, we simulated one thousand truncated distribution of standard errors with the following general characteristics: <!--# Georgina, I have used the Truncated Normal, we can obviously do sensitivity analyses with other ones -->

@@

<!--# Charlotte, I am particularly keen for you to code independently the SE calculation, maybe use all the other parts as they are, but check the SEs. Once you have re-assured yourself about it, could you please read the code for the other steps below, but break it down by taking it out of the function and probably out of the loop and verifying that you would get the same results for one dataset, e.g. the first one in the list of 1000. It would be amazing to verify the results this way. Finally, as discussed, we can vary the parammeters of the distribution if we have better a prioris-->

@@

### Random Effects Metaregression

We estimated the pooled standardised mean difference for each arm by using a random effects metanalysis implemented in R's metafor package. The main underlying assumption of random effects metanalysis is that each study's true effect size $\theta_{k}$ is affected not only by sampling error $\epsilon_{k}$, but also by $\zeta$ which represents heterogeneity between studies, allowing each study's estimate to vary along a distribution of effects, and the distribution of true effect sizes termed $\tau^2$. Therefore, we can estimate a two stage model with:

$$Y_i \sim \mathcal{N}(\theta_i, \sigma_i^2)
$$

@@

$$u_i \sim N(0, \tau^2) $$ {#eq-8}

@@

We used maximum likelihood (ML) to estimate model and applied Hartung-Knapp adjustment to reduce the chance of false positives (NOTE: **Charlotte**cite Ioannides on this).

We present our main results as means of the estimates across simulated datasets, for example, the SMDs of each level of the dummy variable above are means across the simulations.

@@

Studies included in the metaregression.

The data for the studies included in this metanalysis are summarised in Supplementary Table 1 and are also available as a csv dataframe on \[<https://github.com/transatlantic-comppsych/apples_oranges>\]. <!--# Charlotte, this should essentially be our df_long dataset, right? Perhaps, for the puproses of the table we can get rid of some of the columns. It would be good to have it in the way Pim had it-->

@@

# add a new multilevel variable to the simulated data for the regression

@@

# extract coefficients and model characteristics.
condition <- levels(list_df_simulated[[1]]$four_level_var)
aggregate_results_overall <-  aggregate_model_results(list_model_1_meta_reg, condition)
aggregate_results_overall <- cbind(aggregate_results_overall, condition)

@@

### Analysis of all studies

#### Number of studies included

@@

#### Metaregression analysis.

We applied metaregression to obtain the SMDs and confidence intervals of each of the four arms.

As can be seen in Figure 1, there were substantial differences between the four arms of the metanalysis with striking difference being the medication and the psychotherapy control arms. In particular, placebo had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$coef_means, 2)`, 95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$upper_ci, 2)`, whereas psychotherapy controls had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$coef_means, 2)` , 95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$upper_ci, 2)`.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
subtitle_text_all <- "Metanalytic estimates of within-group changes: all studies"
plot_means_all <- plot_means_function(df_mean_coefs_from_sim, subtitle_text_all)
#print(plot_means_all)
pdf(file = "plot_means_all.pdf")
plot_means_all
```

![Figure 1.](plot_means_all.pdf)

Table 1 summarises the statistics of the estimated SMDs. <!--# Charlotte, we need to turn these into nice tables with proper columns, rounding etc. -->

```{r, warning=FALSE, message = FALSE, echo=FALSE}

knitr:: kable(df_mean_coefs_from_sim, caption = "Table 1")
```

In Table 2, we present the regression that tests our hypothesis about differences between medication and psychotherapy control. In particular, in this metaregression model, medication control is the reference category (termed intercept) to which all other categories of the dummy variable, including psychotherapy control, are compared. The strongest difference betwen arms, as judged by the z-value, is between the psychotherapy and medication control with a z-value `r round(aggregate_results_overall[aggregate_results_overall$condition == "psychotherapy_control", ]$z_value, 2)` , which yields a very low p-value ( `r sprintf("%.3e",(1-(pnorm(8.949921))))` ).

```{r, warning=FALSE, message = FALSE, echo=FALSE}
knitr:: kable(aggregate_results_overall, caption = "Table 2")
```

### Sensitivity Analyses

We then conducted a series of sensitivity analyses of our results.

#### Effect of standard errors of the SMDs .

It could be argued that the choice of standard errors of the changes for the calculation of the confidence intervals could have affect the results in one or the other direction. To address such concerns we have simulated 1000 different datasets with SMDs coming from a broad distribution. If standard error distributions were influential, this should show up as substantial variability across simulations. We test this idea in the Figure 1 which displays across the 1000 simulations the z-value of the contrast between medication and psychotherapy control arms (the mean of which we presented in Table 2). As can be seen, the variability in the z-score is minimal and consistently far away from the threshold for significance, i.e. the value of z = 1.645.

@@

stab_sims <- stability_sims %>%
  ggplot(aes(x = n_sims, y = z_value)) +
  geom_point() +
  geom_hline(yintercept = 1.65, linetype = "dotted", color = "red") +
  geom_segment(aes(x = 300, xend = 300, y = 2, yend = 1.65),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  annotate("text", x = 300, y = 2.0, label = "p < 0.05 threshold (values above line significant)", hjust = -0.1, vjust = 0) +
  geom_segment(aes(x = 300, xend = 300, y = 7.5, yend = min(stability_sims$z_value-0.3)),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  annotate("text", x = 300, y = 7.5, label = "simulated z-values",
           hjust = -0.1, vjust = 0) +
  theme_minimal() +
  labs(
    x = "Number of Simulations",
    y = "z-value",
    title = "Stability of the Statistic of the Difference between \nMedication and Psychotherapy Control Arms",
    subtitle = "Results from 1000 simulations with within-group standard errors"
  )

pdf(file = "stab_sims.pdf")
stab_sims
dev.off()
```

![](stab_sims.pdf)

#### Comparison of CBT with fluoxetine and escitalopram

CBT is the most widely studied treatment and, similar to fluoxetine, it has metanalytic evidence supporting its use. Both fluoxetine and escitalopram are approved by the FDA as treatments for adolescent depression. As can be seen in Figure 3, the pattern of results is very similar to that of the overall analyses.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
subtitle_text_cbt <- "Metanalytic estimates of within-group changes: cbt, fluox, escit"
plot_means_cbt <- plot_means_function(coef_and_se_means_cbt_fluox_esc_study , subtitle_text_cbt)
pdf(file = "plot_means_cbt.pdf")
plot_means_cbt
dev.off()
```

![](plot_means_cbt.pdf)

#### Excluding studies with waitlist control arms.

It could be argued that waitlist control arms represent outliers amongst psychotherapy control studies that reduce the estimate of the overall efficacy. We therefore analysed the data excluding waitlist control studies. As can be seen, the pattern of results is very similar to that of the overall analyses.

```{r, warning=FALSE, message = FALSE, echo=FALSE}

subtitle_text_no_wl <- "Metanalytic estimates of within-group changes: wl excluded"
plot_means_wl <- plot_means_function(coef_and_se_means_no_wl, subtitle_text_no_wl)
#print(plot_means_wl)

pdf(file = "plot_means_wl.pdf")
plot_means_wl
dev.off()
```

![](plot_means_wl.pdf)

```         
```
